### 2020-05-19 manuscript/manuscript.rmd
worked on body

Historically, there was little debate about the root of the animal tree of life. Porifera-sister (Fig. 1) was widely accepted though rarely tested. By contrast, there has long been uncertainty about the relationship of Ctenophora to other animals [@Wallberg:2004ws]. The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores rather than sponges are our most distant animal relative. Since then many more studies have been published, with some analyses that find support for Ctenophora-sister, some for Porifera-sister, and some neither (Table 1). As the problem has become better characterized, it has become an interesting test-case for addressing difficult phylogenetic questions, which has attracted methodological interest well beyond investigators concerned with early animal evolution. These studies have taken a wide variety of approaches and results have varied across these different analyses. The extensive technical variation across these studies has been important to advancing our understanding of the sensitivity of these analyses, revealing for example that outgroup and model selection can have a large impact on result. But the larger technical variation has also made it difficult to synthesize these results to understand the details of this sensitivity because so many factors, such as which software tools are used, vary so widely across analyses. This variance is often confounded, for example when a particular model is only available in one software package. To advance this problem it is therefore critical to synthesize published results and explore variation in a standardized framework.

@@

## Variation across published studies

@@

High-throughput sequencing allows investigators interested in the phylogeny of early diverging animals to assemble matrices with hundreds or thousands of genes (Table 1). Surprisingly, gene overlap between studies has rarely been assessed. Though there are many genes with homologs present in most known animal genomes (*e.g.*, animal BUSCO genes [@waterhouse2017busco]), studies of animal phylogeny often relied on a wide variety of different bioinformatic approaches to identifying and selecting genes for their matrices. These take different approaches to sequence comparison, filtering, and orthology assessment, and gene selection can be impacted by which datasets are considered (see supplementary text for more details). As a result, the genes selected for analysis can vary widely between studies.

Ingroup selection. XXX

Outgroup selection (the non-animal taxa used to root the animal tree) can affect the phylogenetic inference of animal evolution (Fig. 1C). Several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology [ @ryan2013genome; @pisani2015genomic]. Three progressively more inclusive clades have often been investigated: Choanomalia (animals plus most closely related Choanoflagellata), Holoza (Choanimalia plus more distantly related holozoans) and Opisthokonta (Holozoa + Fungi). It has been suggested the inclusion of more distant related (e.g. Fungi) outgroup taxa tend to recover Ctenophora-sister, whereas increasing support of Porifera-sister often recovered when more closely related outgroup taxa were used with site-heterogeneous Poisson-CAT models [@pisani2015genomic]. In contrast of outgroup selection, sensitivity to ingroup sampling has received less attention.

## Overview of published analyses
Here, we synthesize data and results from previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister from 14 studies (Table 1). These matrices include all relevant studies published before 2020 for which we could obtain data matrices with gene partitions that described boundaries between genes.

### Matrix taxon composition

Taxa sampling plays a critical role in phylogenetic inference, especially for the taxa whose phylogenetic position is poorly understood. However, several data matrices have relatively skewed taxon sampling, with more than 50% taxa sampled in Bilateria compared to other major animal lineages (Extended Data Fig. 1). These skewed data matrices include two studies constructed from ESTs (Dunn2008 [@Dunn:2008ky] and Hejnol2009 [@hejnol2009assessing]) and one study from whole-genome data (Borrowiec2015 [@borowiec2015extracting]). Importantly, several earlier studies before 2015 only include less than five ctenophore species (Extended Data Fig. 1). The rest data matrices have relatively balanced taxon composition, although the total number of sampled taxon varies. In general, increasing taxon sampling is broadly accepted to improve phylogenetic inference and aid in the placement of the earliest diverging animal lineage. Thus, it should be noted that Whelan2017 [@whelan2017ctenophore] (has the most Ctenophore taxa) and Simion2017 [@simion2017large] (contains the most taxon sampling) have much broader taxon sampling than other matrices.

@@

### Matrix gene sampling are largely different across studies

We first used the Metazoa BUSCO genes [@waterhouse2017busco] to evaluate gene composition from all data matrices and the result is unexpected (Extended Data Fig. 2A). The animal BUSCO dataset contains 978 single-copy orthologs (identified from 64 animal genomes) that are most likely shared by most animal genomes, and are commonly used to assess the completeness of assemblies from genomic data [@waterhouse2017busco]. The BUSCO genes also are frequently used as markers in many large-scale phylogenomic studies (e.g., yeasts [@shen2018tempo], spiders [@fernandez2018phylogenomics]). Thus, it is reasonable to expect that these conserved genes should present at a higher percentage than non-BUSCO genes in most matrices. However, we found that gene partitions from all matrices comprised less than 50% of BUSCO genes, even with several data matrices contain less than 20 percent (Extended Data Fig. 2B). Moreover, we identified number of ribosomal RNA genes in each data matrix (Extended Data Fig. 2C) and found that Philippe2009, Nosenko2013_ribo_14615 and Chang2015 matrices possessed much higher content of ribosomal genes than other matrices (Extended Data Fig. 2D). We also annotated each gene partition against SwissProt [@boeckmann:2003swiss] and Gene Ontology databases, and results are shown in Supplementary Table 1.
Moreover, our network analyses show that surprisingly little overlap of gene samplings across all the data matrices from different studies (Fig. 2). Thus, these results strongly suggested that gene sampling largely varies across studies, indicating that different regions of the genomes are sampled with little overlap among different data matrices.

### Support for Porifera-sister and Ctenophora-sister

@@

## New analyses of published matrices are largely consistent with previous

@@

### Drastic changes of phylogenetic signal between site- homogeneous and heterogeneous models

>>>>

Historically, there was little debate about the root of the animal tree of life. Porifera-sister (Fig. 1) was widely accepted though rarely tested. By contrast, there has long been uncertainty about the relationship of Ctenophora to other animals [@Wallberg:2004ws]. The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores rather than sponges are our most distant animal relative. Since then many more studies have been published, with some analyses that find support for Ctenophora-sister, some for Porifera-sister, and some neither (Table 1). As the problem has become better characterized, it has become an interesting test-case for addressing difficult phylogenetic questions, which has attracted methodological interest well beyond investigators concerned with early animal evolution. These studies have taken a wide variety of approaches and results have varied across these different analyses. The extensive technical variation across these studies has been important to advancing our understanding of the sensitivity of these analyses, revealing for example that outgroup and model selection can have a large impact on result. But the larger technical variation has also made it difficult to synthesize these results to understand the details of this sensitivity because so many factors, such as which software tools are used, vary so widely across analyses. This variance is often confounded, for example when a particular model is only available in one software package.

To advance this problem it is therefore critical to synthesize published results and explore variation in a standardized framework. Here, we synthesize data and results from previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister from 14 studies (Table 1). These matrices include all relevant studies published before 2020 for which we could obtain data matrices with gene partition annotations.

@@

## Variation in models and sampling across published studies

@@

Sampling of ingroup taxa (animals) has always been heavilly biased toward Bilateria, but sampling of non-bilaterian groups, including sponges and ctenophores, has improved over time as studies add additional taxa. As sequencing technologies have developed, there has also been some differences across studies in which members of particular groups are included as older shallower datasets are replaced with new deeper datasets from closely related species. Ingroup taxon sampling across clades varies widely between studies (Extended Data Fig. 1). The first studies that use new sequencing technologies tend to have a greater fraction of species belonging to Bilateria, since it includes many of the best-studied animal groups and all caononical model species where these technologies are first deployed. More than 50% taxa sampled were Bilateria in the earliest transcriptome (Dunn2008 [@Dunn:2008ky] and Hejnol2009 [@hejnol2009assessing]) and genome genome (Borrowiec2015 [@borowiec2015extracting]) studies considered here. Importantly, several earlier studies before 2015 only include less than five ctenophore species (Extended Data Fig. 1). Within each clade, there is often considerable variation in taxon sampling and therefore little species overlap across studies (Extended Data Fig. XXXX).

@@

Sampling of outgroup taxa (non-animals, in this case) is critical to phylogenetic rooting questions, since the node where the outgroup attaches to the ingroup tree is the root of the ingroup. There has therefore been extensive focus on improving outgroup sampling when testing phylogenetic hypotheses about rooting [XX Citation neede]. To explore the issue in the context of rooting animals, several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology [ @ryan2013genome; @pisani2015genomic]. Three progressively more inclusive clades have often been investigated: Choanomalia (animals plus most closely related Choanoflagellata), Holoza (Choanimalia plus more distantly related holozoans) and Opisthokonta (Holozoa + Fungi). It has been suggested the inclusion of more distant related (e.g. Fungi) outgroup taxa tend to recover Ctenophora-sister, whereas increasing support of Porifera-sister often recovered when more closely related outgroup taxa were used with site-heterogeneous Poisson-CAT models [@pisani2015genomic].

High-throughput sequencing allows investigators interested in the phylogeny of early diverging animals to assemble matrices with hundreds or thousands of genes (Table 1). Studies of animal phylogeny often relied on a wide variety of different bioinformatic approaches to identifying and selecting genes for their matrices. These take different approaches to sequence comparison, filtering, and orthology assessment, and gene selection can be impacted by which datasets are considered (see supplementary text for more details). As a result, the genes selected for analysis varies widely between studies (Fig. XX). This reflects the different methods used to collect sequence data and different criteria for gene selection. There are several ways to summarize and assess this variation in gene sampling. We annotated each gene partition against SwissProt [@boeckmann:2003swiss] and Gene Ontology databases (Supplementary Table 1). These annotations show XXX. We identified number of ribosomal RNA genes in each data matrix (Extended Data Fig. 2C) and found that Philippe2009, Nosenko2013_ribo_14615 and Chang2015 matrices possessed much higher content of ribosomal genes than other matrices (Extended Data Fig. 2D). We also identified BUSCO genes [@waterhouse2017busco] in each matrix (Extended Data Fig. 2A). The animal BUSCO dataset contains 978 single-copy orthologs (identified from 64 animal genomes) that are broadly shared by most animal genomes, and are commonly used to assess the completeness of assemblies from genomic data [@waterhouse2017busco]. The BUSCO genes also are frequently used as markers in many large-scale phylogenomic studies (e.g., yeasts [@shen2018tempo], spiders [@fernandez2018phylogenomics]). We found that gene partitions from all matrices comprised less than 50% of BUSCO genes, even with several data matrices contain less than 20% (Extended Data Fig. 2B). This is an indication that gene sampling criterial are not only based on breadth of sampling, but also on other criteria such as features of molecular evolution and sampling effects such as sifferences in expression level.

## Variation in published results

@@

## New standardized analyses of published matrices

@@

### Drastic differences in phylogenetic signal between site- homogeneous and heterogeneous models
